LDpred2: better, faster, stronger

### 2020-04-22 paper/paper-ldpred2.tex
Add algo

To estimate effect sizes $\beta_j$, \cite{vilhjalmsson2015modeling} use a Gibbs sampler [REALLY A GS?]. They compute the probability that variant $j$ is causal as 
\[\bar{p}_j = P\left(\beta_j \sim \mathcal N(\cdot,\cdot) ~|~ \tilde{\beta}_j\right) = \dfrac{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace}{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace + \dfrac{1-p}{\sqrt{\dfrac{1}{N}}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{\tilde{\beta}_j^2}{\dfrac{1}{N}}\right\rbrace}~,\]
which we rewrite as
\begin{equation}
\bar{p}_j = \dfrac{1}{1 + \dfrac{1-p}{p} \sqrt{1 + \dfrac{N h^2}{Mp}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{N \tilde{\beta}_j^2}{1 + \dfrac{Mp}{N h^2}}\right\rbrace}~.
\end{equation}
Computing $\bar{p}_j$ using the second expression is important to avoid numerical issues when $N \tilde{\beta}_j^2$ is large.

Then, $\beta_j$ is sampled according to
\begin{equation}\label{eg:beta_random}
\beta_j ~|~ \tilde{\beta}_j \sim \left\{
\begin{array}{ll}
\mathcal N\left(\dfrac{1}{1 + \dfrac{M p}{N h^2}} \tilde{\beta}_j, \dfrac{1}{1 + \dfrac{M p}{N h^2}}\dfrac{1}{N}\right) & \mbox{with probability $\bar{p}_j$,} \\
0 & \mbox{otherwise.}
\end{array}
\right.
\end{equation}

%\begin{minipage}{.7\textwidth}
\begin{algorithm}[H]
\caption{LDpred} 
\begin{algorithmic}
	\For {$i=1,\ldots,N_{iter}$}
	\For {$j=1,\ldots,M$} \Comment{put some comments here}
	\State Run ... \Comment{put some comments here}
	\State Compute... \Comment{put some comments here}
	\EndFor
	\State Optimize ...
	\State $\theta_{old}\leftarrow\theta$
	\EndFor
\end{algorithmic} 
\end{algorithm}
%\end{minipage}

[TO COMPLETE + CLEARER -> SHOULD SHOW THAT IT IS ITERATIVE]

@@

The first extension consists in estimating $p$ and $h^2$ in the model, as opposed to testing several values of $p$ and estimating $h^2$ using constrained LD score regression \cite[]{bulik2015ld}. This makes LDpred2-auto a method free of hyper-parameters which can therefore directly be applied to data without the need of a validation dataset to choose best-performing hyper-parameters.
To estimate $p$, we count the number of causal variants (i.e.\ $m_c = \sum_j(\beta_j \neq 0)$ in equation \ref{eg:beta_random}). 
We can assume that $m_c \sim \text{Binom}(M, p)$, so if we place a prior $p \sim \text{Beta}(1, 1) \equiv \mathcal{U}(0, 1)$, we can sample $p$ from the posterior $p \sim \text{Beta}(1 + m_c, 1 + M - m_c)$.
Due to complexity reasons, we could not derive a Bayesian estimator of $h^2$. Instead, we estimate $h^2 = \beta^T K \beta$, where $K$ is the correlation matrix.

The second extension, LDpred2-sparse, aims at providing sparse effect size estimates, i.e.\ that some resulting effects are exactly 0.
When the sparse solution is sought and when $\bar{p}_j < p$, we set $\beta_j = 0$.
Note that LDpred2-auto does not have a sparse option, but it is possible to run LDpred2-sparse with the $p$ and $h^2$ estimates from LDpred2-auto.

>>>>

To estimate effect sizes $\beta_j$, \cite{vilhjalmsson2015modeling} use a Gibbs sampler, as described in algorithm \ref{algo:ldpred}.

First, they estimate the residualized marginal effect for variant $j$ as [SHOULD VERIFY]
\begin{equation}\label{eq:beta_res}
\tilde{\beta}_j = \hat{\beta}_j - \boldsymbol{\beta}^T_{\text{-}j} \boldsymbol{K}_{\text{-}j, j}
\end{equation}
where $\boldsymbol{K}_{\text{-}j, j}$ is the $j$-th column without the  $j$-th row of the correlation matrix and $\boldsymbol{\hat{\beta}}$ are the marginal effect sizes.
Then, they compute the probability that variant $j$ is causal as 
\[\bar{p}_j = P\left(\beta_j \sim \mathcal N(\cdot,\cdot) ~|~ \tilde{\beta}_j\right) = \dfrac{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace}{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace + \dfrac{1-p}{\sqrt{\dfrac{1}{N}}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{\tilde{\beta}_j^2}{\dfrac{1}{N}}\right\rbrace}~,\]
which we rewrite as
\begin{equation}
\bar{p}_j = \dfrac{1}{1 + \dfrac{1-p}{p} \sqrt{1 + \dfrac{N h^2}{Mp}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{N \tilde{\beta}_j^2}{1 + \dfrac{Mp}{N h^2}}\right\rbrace}~.\label{eq:postp}
\end{equation}
Computing $\bar{p}_j$ using the second expression is important to avoid numerical issues when $N \tilde{\beta}_j^2$ is large.

Then, $\beta_j$ is sampled according to
\begin{equation}\label{eg:random_beta}
\beta_j ~|~ \tilde{\beta}_j \sim \left\{
\begin{array}{ll}
\mathcal N\left(\dfrac{1}{1 + \dfrac{M p}{N h^2}} \tilde{\beta}_j, \dfrac{1}{1 + \dfrac{M p}{N h^2}}\dfrac{1}{N}\right) & \mbox{with probability $\bar{p}_j$,} \\
0 & \mbox{otherwise.}
\end{array}
\right.
\end{equation}

Therefore, the posterior mean of $\beta_j$ is given by 
\begin{equation}\label{eg:post_beta}
\omega_j = \dfrac{\bar{p}_j \tilde{\beta}_j}{1 + \dfrac{M p}{N h^2}} ~.
\end{equation}

%\begin{minipage}{.7\textwidth}
\begin{algorithm}[H]
\caption{LDpred, with hyper-parameters $p$ and $h^2$}\label{algo:ldpred}
\begin{algorithmic}[1]
	\State $\boldsymbol{\hat{\beta}} \gets \dfrac{\boldsymbol{\hat{\gamma}}}{\boldsymbol{\text{sd}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}}$ \Comment{See previous section}
	\State $\boldsymbol{\Omega} \gets \boldsymbol{0}$
	\For {$k=1,\ldots,N_\text{burn-in} + N_\text{iter}$} \Comment{Gibbs iterations}
	\For {each variant $j$} \Comment{All variants}
	\State Compute $\tilde{\beta}_j$ according to \eqref{eq:beta_res}
	\State Compute $\bar{p}_j$ according to \eqref{eq:postp}
	\State Sample $\beta_j$ according to \eqref{eg:random_beta}
	\State Compute $\omega_j$ according to \eqref{eg:post_beta}
	\EndFor
	\State Compute $\boldsymbol{\beta}^T \boldsymbol{K} \boldsymbol{\beta}$ (estimate of $h^2$) to check divergence \Comment{Return 0s if divergence}
	\If {$k > N_\text{burn-in}$} 
	\State Increment $\boldsymbol{\Omega}$ with $\boldsymbol{\omega}$ 
	\EndIf
	\EndFor
	\State $\boldsymbol{\Omega} \gets \boldsymbol{\Omega} /  N_\text{iter}$
	\State Return $\boldsymbol{\Omega} \cdot  \boldsymbol{\text{sd}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}$ \Comment{See previous section}
\end{algorithmic} 
\end{algorithm}
%\end{minipage}

[CHANGE NOTATION OF SUMSTATS / SD -> SE?]

@@

The first extension consists in estimating $p$ and $h^2$ in the model, as opposed to testing several values of $p$ and estimating $h^2$ using constrained LD score regression \cite[]{bulik2015ld}. This makes LDpred2-auto a method free of hyper-parameters which can therefore directly be applied to data without the need of a validation dataset to choose best-performing hyper-parameters.
To estimate $p$, we count the number of causal variants (i.e.\ $m_c = \sum_j(\beta_j \neq 0)$ in equation \ref{eg:random_beta}). 
We can assume that $m_c \sim \text{Binom}(M, p)$, so if we place a prior $p \sim \text{Beta}(1, 1) \equiv \mathcal{U}(0, 1)$, we can sample $p$ from the posterior $p \sim \text{Beta}(1 + m_c, 1 + M - m_c)$.
Due to complexity reasons, we could not derive a Bayesian estimator of $h^2$. Instead, we estimate $h^2 = \beta^T K \beta$, where $K$ is the correlation matrix.
These parameters $p$ and $h^2$ are updated after the inner loop in algorithm \ref{algo:ldpred}, then these new values are used in the next iteration of the outer loop.

The second extension, LDpred2-sparse, aims at providing sparse effect size estimates, i.e.\ that some resulting effects are exactly 0.
When the sparse solution is sought and when $\bar{p}_j < p$, we set $\beta_j$ and $\omega_j$ to 0 (lines 6-8 of algorithm \ref{algo:ldpred}).
Note that LDpred2-auto does not have a sparse option, but it is possible to run LDpred2-sparse with the $p$ and $h^2$ estimates from LDpred2-auto.
